Automatic Amortized Analysis for Imperative Code

### 2014-03-05 main.tex
improve notations
bug in AssignOpK
add simple cost aware semantics

We start with a programming language with loops counting
downwards to 0.  Statements and expressions are defined
below.  Assignments of complex expressions can be simulated
by sequencing assignments of simple expressions.
%
\begin{mathpar}
S := \code{skip}
\mid \code{assert}(\Gamma)
\mid x \gets E
\mid \code{while}(0 < x)~S
\mid \code{if}(?)~S~\code{else}~S
\mid S;S

@@

\section{Decreasing differences}
The second iteration is a system more symmetric where the potentials
are linear functions in differences of variables.  From now on $k$
denotes a non-negative constant.
$$
S := \code{while}(k < x - y)~S \mid \dots
$$

@@

\Rule{AssignOpK}
{  \Gamma \models \Gamma'[x/x \pm y], y = k
\\ \Gamma \models x \pm y \in [x, u]
\\ \textstyle q'_y = q_y + \sum_u q^1_{xu} - \sum_v q_{xv}
\\ q_{xu} = q^1_{xu} + q^2_{xu}
}
{ \textstyle
  \Gamma; Q \vdash x \gets x \pm y \dashv \sum_u k \cdot q^2_{xu} + Q'; \Gamma'
}

\and \Rule{AssignOp}
{  \Gamma \models \Gamma'[x/x \pm y]
\\ \Gamma \models x \pm y \in [x, u]
\\ \textstyle q'_y = q_y + \sum_u q_{xu} - \sum_v q_{xv}
}
{ \textstyle
  \Gamma; Q \vdash x \gets x \pm y \dashv Q'; \Gamma'
}

@@

>>>>

We start with a programming language with loops counting
downwards to 0.  Statements and expressions are defined
below.  Assignments of complex expressions can be simulated
by sequencing assignments of simple expressions.
%
\begin{mathpar}

S := \code{skip}
\mid \code{assert}(\Gamma)
\mid x \gets E
\mid \code{while}(0 < x)~S
\mid \code{if}(?)~S~\code{else}~S
\mid S;S

@@

\section{Decreasing differences}
The second iteration is a system more symmetric where the potentials
are linear functions in differences of variables.  From now on $k$
denotes a non-negative constant.
$$
S := \code{while}(k < x - y)~S \mid \dots
$$
%
In the rules below we take as convention that $q_{x}$ for a variable
$x$ is in fact $q_{x0}$ where 0 is a special variable fixed to zero
in all contraint contexts.

@@

\Rule{AssignOpK}
{  \Gamma \models \Gamma'[x/x \pm y]
\\ \Gamma \models |y| = k
\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
\\ \textstyle q'_y = q_y
      + \sum_{u \in \mathcal U} p_{xu}
      - \sum_{v \not\in \mathcal U} q_{xv}
\\ q_{xu} = p_{xu} + r_{xu}
}
{ \textstyle
  \Gamma; Q
  \vdash x \gets x \pm y \dashv
  \sum_u k \cdot r_{xu} + Q'; \Gamma'
  % q_x is not omitted, v_{x} is in fact a shorthand for
  % v_{x0} where 0 is a variable equal to 0.
}

\and \Rule{AssignOp}
{  \Gamma \models \Gamma'[x/x \pm y]
\\ \mathcal U = \left\{ u \mid \Gamma \models x \pm y \in \inter x u \right\}
\\ \textstyle q'_y = q_y
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  \Gamma; Q \vdash x \gets x \pm y \dashv Q'; \Gamma'
}

@@

\section{Cost aware semantics}

This big step semantics counts the number of iterations of
while loops in a program as it executes it.  Non-deterministically,
the {\sc S:Bubble} rule can be called to interrupt a computation.
It has the same effect as raising an exception.  We symbolize an
immutable heap with the notation $\circ H$, this kind of heap just
bubbles up through the semantics. In the rules, $X$ stands for
either $H$ or $\circ H$ for some $H$.

\Rule{S:Bubble}
{}
{ (H, S) \Downarrow_0 \circ H }

\and \Rule{S:Skip}
{}
{ (H, \code{skip}) \Downarrow_0 H }

\and \Rule{S:Assign}
{}
{ (H, x \gets e) \Downarrow_0 H[x \gets \llbracket e \rrbracket_H] }

\and \Rule{S:Assert}
{ \Gamma \models H }
{ (H, \code{assert}(\Gamma)) \Downarrow_0 H }

\and \Rule{S:WhileStop}
{ \neg (k < \llbracket x - y \rrbracket_H) }
{ (H, \code{while}(k < x - y)~S) \Downarrow_0 H }

\and \Rule{S:WhileStep}
{  k < \llbracket x - y \rrbracket_H
\\ (H, S; \code{while}(k < x - y)~S) \Downarrow_n X
}
{ (H, \code{while}(k < x - y)~S) \Downarrow_{n+1} X }

\and \Rule{S:If}
{  (H, S_i) \Downarrow_n X
\\ i \in \{1, 2\}
}
{ (H, \code{if}(?)~S_1~\code{else}~S_2) \Downarrow_n X }

\and \Rule{S:Seq}
{  (H, S_1) \Downarrow_{n_1} H'
\\ (H', S_2) \Downarrow_{n_2} H''
}
{ (H, S_1; S_2) \Downarrow_{n_1 + n_2} H'' }

\and \Rule{S:SeqBubble}
{ (H, S_1) \Downarrow_n \circ H' }
{ (H, S_1; S_2) \Downarrow_n \circ H' }
